---
title: "Bread dough experiment analysis"
output: html_document
---

```{r setup, include=FALSE}
require(daewr)
knitr::opts_chunk$set(echo = TRUE)
```

### Building and analyzing Completely Random Design (CRD) linear model with bread dough experiment

1. Gather some R packages 

```{r package}
## R package from the book
require(daewr)
## gmodels package that returns estimable contrasts
require(gmodels)
## nice visualization and summary tools for models, optional
require(jtools)


```

2. Check out the `bread` dataset. Note: only print the whole dataset in your write-up when you know the dataset is small, otherwise, something like `print(head(<your data>))` is more visually friendly.

```{r data}
## 
print(bread)

```

3. Build model using the R function `lm()`: treatment = time to rise, response variable = height of dough. Use `?lm` in R console to learn more about the function.
```{r model}

mod0 <- lm(height ~ time, data = bread)

## take a look at summary stats to see effects estimates
summary( mod0 )

## you can also use the summ function from jtools for a more neat representation of the result

#summ( mod0 )
```


4. *Parameter estimation* : comparing the LS method with the result from `lm()`

Let's start by creating the necessary matrices
```{r verify_est}
# y = response variable
y <- bread$height
# X = design matrix
X <- matrix(nrow = 12,ncol = 3)
X[,1] <- rep(1,12)
X[,2] <- c(rep(0,4),rep(1,4),rep(0,4))
X[,3] <- c(rep(0,8),rep(1,4))

XtX <- t(X)%*%X

XtX.inv <- solve(XtX)

beta.hat <- XtX.inv%*%t(X)%*%y

print(beta.hat)

print(coef(mod0))
```

5. Just to verify - if we include the vector of ones as noted in the textbook to create a 12 x 4 design matrix, $X$, instead, what would happen to the $(X'X)^{-1}$ ?

```{r check_matrix}
# Z = alternative design matrix
Z <- matrix(nrow = 12,ncol = 4)
Z[,1] <- rep(1,12)
Z[,2] <- c(rep(1,4),rep(0,8))
Z[,3] <- c(rep(0,4),rep(1,4),rep(0,4))
Z[,4] <- c(rep(0,8),rep(1,4))

ZtZ <- t(Z)%*%Z

```

6. Alternatively, we can estimate the contrasts between effects $\tau_i$

```{r contrasts}

t1.t2<-fit.contrast(model = mod0, 'time', c(1,-1,0))
t1.t3<-fit.contrast(model = mod0, 'time', c(1,0,-1))
```


7. *Hypothesis testing* , we now want to test the hypothesis $H_0 : \mu_1 = \mu_2 = \cdots = \mu_t$ vs the alternative, which is unequal effect for any pair. F test looks at the ratio of variation from two different sources. In our case, we are interested in treatment variation vs model variation. In R, we can use `lm()` to do ANOVA, but `aov` is a simple way that is sometimes preferred. 

```{r hyp_test}
mod1 <- aov(height~time, data = bread)
print(summary(mod1))
```

In the F-test above, assuming a significance level $\alpha = 0.5$, we should reject $H_0.$


8. *Model assumption check* - how valid are the results of our analysis so far? One way to find out is the verify that the model assumptions are met, which we can show in R using the following commands:

```{r model_assumptions}
# generate 2 by 2 canvas for plots
par( mfrow = c(2,2))
plot(mod0, which = 5)
plot(mod0, which = 1)
plot(mod0, which = 2)
plot(residuals(mod0)~loaf, main = 'Residuals vs Exp. Unit', data = bread)
```